# Implements a linear Kalman filter.

$$ x_{t+1} = Ax_t + \epsilon_t $$ $$ y_t = H  x_t+ \nu_t $$ $$ \epsilon ~ N(0,Q)
$$ $$ \nu ~ N(0,R) $$

## Staionery

$$ x_{t+1} = t + \epsilon_t$$ $$ y_t = x_t + \nu_t $$


```{r} 
Kal = list(A = matrix(1,1,1), B = matrix(0,1,1), H = matrix(1,1,1), x_hat
= matrix(3,1,1), P_hat = matrix(1,1,1), Q = matrix(0.00001,1,1), R =
matrix(0.1,1,1))
```

Kalman interativ procedyre: One step

```{r} 
step = function(Kal,measured, control_vector=0){
  with(Kal,{ 
   #-----------------Prediction step----------------------------- 
    predicted_state_estimate = A %*% x_hat + B %*% control_vector 
    predicted_prob_estimate = (A %*% P_hat) %*% t(A) + Q 
    #-------------------Observation step--------------------------- 
    innovation = measured - H %*% predicted_state_estimate 
    innovation_covariance = H %*% predicted_prob_estimate%*%t(H) + R
    #-------------------Update step------------------------------- 
    kalman_gain = predicted_prob_estimate %*% t(H) %*% solve(innovation_covariance) 
    current_state_estimate = predicted_state_estimate + kalman_gain %*% innovation
    current_prob_estimate = (diag(1,1)-kalman_gain %*% H) %*% predicted_prob_estimate
    Kal$x_hat = current_state_estimate 
    Kal$P_hat = current_prob_estimate 
    return(Kal) 
  }) 
}
```

Simulation of voltage mesurement

```{r} 
getVoltage = function() rnorm(1,mean=1.25,sd=0.25) 
```


simulat 60 time step of mesurement and prediction
```{r} 

n=60

dat = data.frame(messuredvotage=rep(NA,n),kalman=rep(NA,n))
for (i in 1:n){ 
  measured = getVoltage() 
  dat[i,] = c(matrix(measured,1,1),Kal$x_hat) 
  Kal = step(Kal,measured) 
} 
```

Plot the out put

```{r} 
plot(dat$kalman, type='l', ylim = range(c(dat$kalman,dat$messuredvotage)), col='blue', lwd=2) 
points(dat$messuredvotage,col='red') 
abline(h=1.25, col='green') 
```


## Time varing regresion


```{r, echo=FALSE} 
t=0:200 
a = 1 - t/100 
x=t %% 12 + t/100 + rnorm(length(t),sd = 0.5) 
y= (x + rnorm(length(x),sd=0.2))*a/5 

```

We have a timevaring x value (a value we know/control)

```{r} 
plot(t,x,type='l') 
```

and a y that somhow depend on x

```{r} 
plot(t,y,type='l')

``` 
There is a coreelation between x and y:

```{r}
plot(x,y) 
abline(lm(y~x)) 
```

A simpel regresion may not be the right thing:

```{r} 
m = lm(y~x) 
summary(m) 
plot(t,y,col='blue') 
points(t,predict(m),type='l',col = 'red')
```

The Kalmnan filter can be used as iterative updating the raltion between $x$ and
$y$. Here the stae varible is the coeficient vector:

$$ y_t = [1 x_t] \beta_t + \nu $$
$$ \beta_{1+t} = I(1) \beta_t + \epsilon $$

The step function then have to chnged so the $H$ can vary with time:

```{r}
step = function(Kal, measured, x, control_vector=0){
  H = matrix(c(1, x),1,2) 
  with(Kal,{
    #-----------------Prediction step----------------------------- 
    predicted_state_estimate = A %*% x_hat + B %*% control_vector 
    predicted_prob_estimate = (A %*% P_hat) %*% t(A) + Q 
    #-------------------Observation step--------------------------- 
    innovation = measured - H %*% predicted_state_estimate 
    innovation_covariance = H %*% predicted_prob_estimate%*%t(H) + R 
    #-------------------Updatestep------------------------------- 
    kalman_gain = predicted_prob_estimate %*% t(H) %*% solve(innovation_covariance) 
    current_state_estimate = predicted_state_estimate + kalman_gain %*% innovation 
    current_prob_estimate = (diag(1,2)-kalman_gain %*% H) %*% predicted_prob_estimate 
    Kal$x_hat = current_state_estimate 
    Kal$P_hat = current_prob_estimate 
    return(Kal) 
  })
}
```

And the kaman list is:

```{r}
Kal = list(A = diag(c(1,1)), 
           B = matrix(0,2,1), 
           x_hat = matrix(1,2,1), 
           P_hat =diag(1,2), 
           Q = diag(c(0.01,0.01)), 
           R = matrix(0.1,1,1))
```

So vi do the iterations:

```{r}
n=length(x)
kalman=data.frame(a=rep(NA,n), b=rep(NA,n), y_hat=rep(NA,n))
for (i in 1:n){ 
  kalman[i,1:2] = t(Kal$x_hat) 
  kalman[i,3] = matrix(c(1, x[i]),1,2) %*% Kal$x_hat 
  Kal = step(Kal,y[i],x[i])

}

```

```{r}
plot(t,y,col='blue') 
points(t,predict(m),type='l',col = 'red') 
points(t,kalman$y_hat,type='l',col='green',lwd=2)
plot(t,kalman$a, type='l',col='blue')
points(t,kalman$b, type='l',col='red') 
abline(h=coef(m)[1],col='red',lty=2) 
abline(h=coef(m)[2],col='blue',lty=2) 
abline(h=0,col='grey')
```

## Libraries

There is different librayes for doing kalman filtering eg KFAS
Unfortinly y´they use other symbols in the equations and i am not sure I have interpreted evry thing right:

```{r}
library(KFAS)

Yt = matrix(y,n,1) 
Zt = array(rbind(rep(1,n),x),c(1,2,n)) # H
Qt = diag(c(0.01,0.01)) 
Rt = matrix(1,2,2) 
Ht = matrix(0.1,1,1) 
Tt = diag(c(1,1)) 
km = SSModel(y~ -1 + SSMcustom(Z=Zt, R=Rt, T=Tt, Q=Qt)) 
```


```{r}
plot(t,y,col='blue') 
points(t,predict(m),type='l',col = 'red') 
#points(t,kalman$y_hat,col='green',lwd=2)
points(t,predict(km),col='green', type='l',lwd=2)
```


